---
title: "Statanalysis User Manual"
author: "Sophia Arias"
date: "2024-12-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Statanalysis

The R Package statanalysis was created for a bioinformatics software development course (BI503G). The main goal of the package is the use the function statanalysis() and have it run. While running it will read in two files: count data file and sample data file.

After the files are read and the data is stored in variables, filtering and statistical analysis will start being performed. DEGs are sought out and stored in a variable and written to an excel file for later use. The DEGs are then used further in enrichment analysis where enrichGO and enrichKEGG will be used. 

After that, dot plots, tree plots, and cnet plots are made and saved in .png format.

## Preparing the space

To be able to use the R package *statanalysis*, it is recommended that you have the following packages already downloaded (the function will load them all):

* devtools
* edgeR
* clusterProfiler
* enrichplot
* org.Hs.eg.db
* openxlsx
* DESeq2
* AnnotationDbi
* ggplot2

```{r Installing, eval = F, echo = T}
install.packages(c("openxlsx", "ggplot2")

##Install Bioconductor packages
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("edgeR", "clusterProfiler", "enrichplot", "org.Hs.eg.db", "DESeq2", "AnnotationDbi"))

#Load the required libraries
install.packages("devtools")
```


## After Downloading Packages

Once the packages have been downloaded, the R package can be downloaded as well:
```{r Downloading, eval=F, echo=T}
library(devtools)
install_github("imszofko/statanalysis");
```


The name of the package is Assignment2StatAnalysis, you will load this up by using the function library().

Once the package is loaded, there are only a few packages to use, this is the order they should be used (see code below).  **Both data file names should be written in quotes.**

```{r Using, eval=F, echo=T}
library(Assignment2StatAnalysis)

countdata <- countData("E-MTAB-2523.counts.txt")

sampledata <- sampleData("E-MTAB-2523_sample table.txt")

filterdata <- filtering(countdata)

stats <- statanalysis(countdata, sampledata)

linearmodel <- linmodel(countdata, sampledata)

```

After the functions run, it will save data files and .png of the enrichment analysis visuals in the user's current working directory.